Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-05-25 paper.tex
More updates

% \begin{itemize} \item A potential source of error in sc2ts ARGs comes from
% incorrectly reported collection dates of the samples, which can cause sample
% genomes to be attached to the wrong parts of the ARGs. Thus far, we have
% attempted to correct the most offending errors in the ARGs by filtering out
% ``time travellers'', but the sampling time can still be wrong by days or tens
% of days. ***describe possible approaches to correct this?*** \item To avoid
% time travellers, we used a maximum submission delay, but this will
% systematically bias against submissions from countries with greater submission
% delays (Kalia et al.). An area of future work is to add back the excluded
% sequences to a built ARG to assess which ones are likely true time travellers.
% \item We have used a single mismatch value in all the HMM runs, set so that
% having three mutations is as likely as having a single recombination. This
% parameter choice is arbitrary, and higher values may reduce the number of
% origins for those Pango lineages that were inferred to descend from more than
% one recombinant ancestor. In addition, because both the mutation rate and
% recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
% allow the mismatch value to vary along the genome. \item By treating mutations
% equally in the current HMM implementation, we implicitly assume that the rates
% of different types of mutations are identical. However, as reported previously
% \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in
% SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide
% substitution models (e.g., allowing for different rates of transition and
% transversion) can be incorporated into HMM likelihood calculations. \item Our
% current masking strategy handles indels as missing data. Doing this can remove
% information helpful to detect recombinants. For example, sequences of the Pango
% recombinant XP were not identified as recombinants, because XP’s characteristic
% multibase deletion was masked. \item Daily trees are built using the UPGMA
% algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of
% the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm
% to build daily trees would leave these "soft polytomies" in, removing the need
% for the un-resolving step. \end{itemize}

@@

\subsection{Tree inference from HMM daily sample clusters}
\label{sec:sample-cluster-tree-inference}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed below
in Section XXX.

With tens of thousands of samples being added to the ARG per day,
we can have clusters of hundreds of sequences attaching to the same node.
While some of these samples will require no extra mutations
(because they are identical to the attachment node), in general there
will be complex patterns of shared mutations among the samples
reflecting their evolutionary relationships. A natural way to infer
these within-cluster evolutionary relationships is to use standard
tree-building algorithms.
We can infer a likely tree relating the
samples in a cluster independently of the other samples in a
daily batch, and then attach the tree (and mutations)
to the ARG at the node identified by the HMM.

We currently use the implementation of the UPGMA algorithm~\citep{Michener1957-tr}
from SciPy~\citep{Pauli2020-scipy} to build trees from sample
clusters, and map mutations back to this topology using maximum parsimony.
This approach was chosen mainly for simplicity and because of the
speed and reliability of the implementation.
An issue with the UPGMA algorithm is that it generates a strictly
binary tree, creating internal nodes
supported by no informative site (i.e., having no mutation immediately
ancestral to them). To avoid such false precision we post-process
to remove unsupported internal nodes, representing the relationship
between $k$ identical descendants of the node as a polytomy of size $k$.

There are well-known issues with using such a simple algorithm for inferring
evolutionary relationships~\citep{Felsenstein2004-inferring}.
Since this within-cluster tree building has a significant influence on the
overall ARG topology (see Results), implementing more sophisticated
tree building methods which keep track of the location of the required mutations
(rather then inferring post-hoc by parsimony), is a likely avenue
for improvements in overall inference quality.

\subsection{Parsimony-increasing heuristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants, see the
next section) under the HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is then to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing''.

Mutation collapsing inspects the siblings of a newly attached node
to check if any of them share (a subset of) the mutations that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations, and make that new node the parent of the
siblings carrying those shared mutations. The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse. [TODO back ref to the
Results section to discuss how important these nodes are]

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occured on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent which carries all its mutations except those
causing the reversions on the newly added node.
[TODO back ref to the
Results section to discuss how important these nodes are]

Together these operations substantially increase the overall parsimony
of the inferred ARGs, most noticably in terms of the number of
reversion mutations inferred. However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Since the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual, it is clear that there
is room for improvement, and that further parsimony-increasing
heuristics would likely be of benefit.

\subsection{Treatment of recombinants}
\label{sec:treatment_recombinants}
A sample sequence is designated as a recombinant if the most likely
path inferred by the LS HMM for that sample contains at least
one switch between parents. Recombinant sequences are mostly treated
identically to non-recombinants, as we simply need to reason about
a path of parent nodes along genome intervals rather than a
single parent over the whole genome, which is naturally handled by
the succinct tree sequence data structure and \texttt{tskit} library
(see the Ancestral Recombination Graphs section above). To facilitate
analysis and to help understand the robustness of recombinants
we perform some additional steps in \texttt{sc2ts}.

The LS HMM may infer identical paths and patterns of mutations
for multiple samples in a daily batch, and so we
create a ``recombination`` node (marked with a specific ``flags'' value)
for each distinct recombinant. (This node is not strictly necessary
but makes it convenient to find recombinants for subsequent analysis.)
Variation within a cluster
of recombinant sequences is handled in the same way as non-recombinants
(see Section~\ref{sec:sample-cluster-tree-inference} above for details).

When the Viterbi algorithm implementation used by the LS HMM infers
recombinant ancestry for a given genome, the point at which inheritance
switches from one parent node to another is the last possible
position.
Thus, the breakpoints used in the ARG represent the rightmost end of the
interval, within which the recombination event could have occurred. To obtain
the leftmost end of this interval, we rerun the HMM using mirrored coordinates
(i.e., from the 3’ end to the 5’ end) and store the information as
metadata on the recombination node.
These left- and rightmost positions
are the recombination intervals reported throughout.
Ultimately all the recombination intervals mean is that the parent sequences
are identical within that interval, and there is therefore no information
about where the recombination occurred. There are many different processes
that can contribute to the size of these intervals, and therefore
understanding how these processes interact to produce wider or narrower
intervals is an important avenue for future work.

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterisation of using
a single ``mismatch ratio'' is very simplistic (see section XXX), and likely
results in a very flat likelihood space where many sequences have equal
likelihood of being chosen. More sophisticated parameterisations where we
condition on known per-site mutation rates (and perhaps per-base subsitution
rates) may lead to better results, although perhaps at the cost of some
reduction in interpretability. Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are chosen
from within the possible interval, rather than the current approach of taking
the rightmost value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events, or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability identify recombinants and to filter poor quality data.

>>>>

% \begin{itemize} \item A potential source of error in sc2ts ARGs comes from
% incorrectly reported collection dates of the samples, which can cause sample
% genomes to be attached to the wrong parts of the ARGs. Thus far, we have
% attempted to correct the most offending errors in the ARGs by filtering out
% ``time travellers'', but the sampling time can still be wrong by days or tens
% of days. ***describe possible approaches to correct this?*** \item To avoid

% time travellers, we used a maximum submission delay, but this will
% systematically bias against submissions from countries with greater submission
% delays (Kalia et al.). An area of future work is to add back the excluded
% sequences to a built ARG to assess which ones are likely true time travellers.
% \item We have used a single mismatch value in all the HMM runs, set so that
% having three mutations is as likely as having a single recombination. This
% parameter choice is arbitrary, and higher values may reduce the number of
% origins for those Pango lineages that were inferred to descend from more than
% one recombinant ancestor. In addition, because both the mutation rate and
% recombination rate varies across the SARS-CoV-2 genome, it may be helpful to
% allow the mismatch value to vary along the genome. \item By treating mutations
% equally in the current HMM implementation, we implicitly assume that the rates
% of different types of mutations are identical. However, as reported previously
% \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in
% SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide
% substitution models (e.g., allowing for different rates of transition and
% transversion) can be incorporated into HMM likelihood calculations. \item Our
% current masking strategy handles indels as missing data. Doing this can remove
% information helpful to detect recombinants. For example, sequences of the Pango
% recombinant XP were not identified as recombinants, because XP’s characteristic
% multibase deletion was masked. \item Daily trees are built using the UPGMA
% algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of
% the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm
% to build daily trees would leave these "soft polytomies" in, removing the need
% for the un-resolving step. \end{itemize}

@@

\subsection{Tree inference from HMM daily sample clusters}
\label{sec:sample-cluster-tree-inference}
For a given daily batch of sample sequences that are to be attached to the
ARG, the LS HMM provides the most likely copying path and required mutations
for each of those sequences. We could directly attach these samples to
the ARG according to the copying paths and sets of mutations, but
such an approach would be highly unparsimonious because there are
often many samples that share the same copying path, with
complex patterns of shared mutations among those clusters of samples.
In this section we discuss the strategy used to infer likely tree
topologies within these clusters of samples sharing the same copying path.
In the majority of cases the copying path will consist of a single
node (i.e., no recombination), and so for simplicity we will assume in the
following discussion that all of the samples in a HMM-defined cluster
attach to a single node. The special treatment afforded to
newly-detected putative recombinants are discussed
in Section~\ref{sec:treatment_recombinants}.

With tens of thousands of samples being added to the ARG per day,
there are often clusters of hundreds of sequences attaching to the same node.
While some of these samples will require no extra mutations
(because they are identical to the attachment node), in general there
will be complex patterns of shared mutations among the samples
reflecting their evolutionary relationships. A natural way to infer
these within-cluster evolutionary relationships is to use standard
tree-building algorithms.
We can infer a likely tree relating the
samples in a cluster independently of the other samples in a
daily batch, and then attach the tree (and mutations)
to the ARG at the node identified by the HMM.

We currently use the UPGMA algorithm~\citep{Michener1957-tr}
as implemented in SciPy~\citep{Pauli2020-scipy} to build trees from sample
clusters, and map mutations back to this topology using maximum parsimony.
This approach was chosen mainly for simplicity and because of the
speed and reliability of the implementation.
An issue with the UPGMA algorithm is that it generates a strictly
binary tree, creating internal nodes
supported by no informative site (i.e., having no mutation immediately
ancestral to them). We avoid such false precision by post-processing
to remove unsupported internal nodes, representing the relationship
between $k$ identical descendants of a node as a polytomy of size $k$.

There are well-known issues with using such a simple algorithm for inferring
evolutionary relationships~\citep{Felsenstein2004-inferring}.
Table~\ref{tab:args} show that
this within-cluster tree building has a significant influence on the
overall ARG topology (Wide ARG: 7.05\% of nodes and parent of
32.27\% of edges; Long ARG: 6.61\% and 19.97\%, respectively),
and therefore applying more sophisticated
tree building methods which keep track of the required mutations
(rather then inferring post-hoc by parsimony), is a likely avenue
for improvements in overall inference quality.

\subsection{Parsimony-increasing heuristics}
\label{sec:parsimony-heuristics}
Attaching trees built from the clusters of samples that copy from
a particular node (or path of nodes for recombinants,
see Section~\ref{sec:treatment_recombinants}) under the
HMM is an inherently greedy strategy,
and can produce inferences that are clearly unparsimonious.
The final step in adding a daily batch of samples to the ARG
is therefore to perform some local updates that target specific
types of parsimony violation in the just-updated regions of the
ARG. There are currently two parsimony-increasing operations
applied, which we refer to as ``mutation collapsing'' and ``reversion
pushing'' (Figure~\ref{fig:overview_sc2ts}D, E).

Mutation collapsing inspects the siblings of a newly attached node
to check if any of them share (a subset of) the mutations that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations, and make that new node the parent of the
siblings carrying those shared mutations. The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse.
Table~\ref{tab:args} shows that
while these nodes constitute a small fraction of the
overall total (Wide ARG: 2.77\%, Long ARG: 4.82\%),
they are parental to a much larger fraction of the
edges (Wide ARG: 12.77\%, Long ARG: 20.47\%).

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occured on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent which carries all its mutations except those
causing the reversions on the newly added node.
Table~\ref{tab:args} shows that
these nodes constitute a similar fraction of the
overall numbers of nodes (Wide ARG: 2.79\%, Long ARG: 4.39\%)
and edges (Wide ARG: 12.66\%, Long ARG: 18.46\%)
as mutation collapsing.

Together these operations substantially increase the overall parsimony
of the inferred ARGs, most noticably in terms of the number of
reversion mutations inferred. These nodes are also being chosen by the
LS HMM as likely choices of parent (as indicated by the node and edge
counts quoted above), also demostrating that the hueristics are successfully
capturing features of real sequences.
However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Since the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual, it is clear that there
is room for improvement, and that further parsimony-increasing
heuristics would likely be of benefit.

\subsection{Treatment of recombinants}
\label{sec:treatment_recombinants}
A sample sequence is designated as a recombinant if the most likely
path inferred by the LS HMM for that sample contains at least
one switch between parents. Recombinant sequences are mostly treated
identically to non-recombinants, as we simply need to reason about
a path of parent nodes along genome intervals rather than a
single parent over the whole genome, which is naturally handled by
the succinct tree sequence data structure and \texttt{tskit} library
(Section~\ref{sec:args}). To facilitate
analysis and to help understand the robustness of recombinants
we perform some additional steps in \texttt{sc2ts}.

The LS HMM may infer identical paths and patterns of mutations
for multiple samples in a daily batch, and so we
create a ``recombination`` node (marked with a specific ``flags'' value)
for each distinct recombinant. (This node is not strictly necessary
but makes it convenient to find recombinants for subsequent analysis.)
Variation within a cluster
of recombinant sequences is handled in the same way as non-recombinants
(see Section~\ref{sec:sample-cluster-tree-inference} above for details).
When the Viterbi algorithm implementation used by the LS HMM infers
recombinant ancestry for a given genome, the point at which inheritance
switches from one parent node to another is the last possible
position. The left-most extent of the breakpoint interval is derived
by sequence comparison between the parents, as described in
Section~\ref{sec:breakpoint_intervals}.

% Thus, the breakpoints used in the ARG represent the rightmost end of the
% interval, within which the recombination event could have occurred. To obtain
% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
% (i.e., from the 3’ end to the 5’ end) and store the information as
% metadata on the recombination node.
% These left- and rightmost positions
% are the recombination intervals reported throughout.
% Ultimately all the recombination intervals mean is that the parent sequences
% are identical within that interval, and there is therefore no information
% about where the recombination occurred. There are many different processes
% that can contribute to the size of these intervals, and therefore
% understanding how these processes interact to produce wider or narrower
% intervals is an important avenue for future work.

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterisation of using
a single ``mismatch ratio'' is very simplistic (Section~\ref{sec:ls}), and likely
results in a flat likelihood space where many recombinant paths have equal
probability of being chosen.
More sophisticated parameterisations where we
condition on known per-site mutation rates (and perhaps per-base subsitution
rates) may lead to better results, although perhaps at the cost of some
reduction in interpretability. Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are chosen
from within the possible interval, rather than the current approach of taking
the rightmost value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events, or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability identify recombinants and to filter poor quality data.
